Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-07-01 manuscript/manuscript.Rmd
Restructuring for word counts and adding untargeted area discussion

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE)
library(tidyverse)
library(RaMS)
library(vegan)
library(cowplot)
library(viridis)
library(scales)
library(ggnewscale)
library(egg)
library(broom)
library(lme4)
library(openxlsx)
library(readxl)
library(ggh4x)
library(ggtext)
library(gridExtra)
library(openxlsx)

@@

This shift in metabolite abundance was largely driven by a corresponding shift in biomass. Particulate carbon and summed metabolite concentration were tightly correlated across all samples and depths ($\beta = 24.9\pm1.82$ (SE), Pearson's $r$ = .813, p-value < 0.001, Supplemental Figure 2). This overall trend was largely driven by higher values at 15 meters and the DCM than at depth, but when analyzed at each depth individually the general correspondence held (Supplemental Figure 2). The DCM and 175 meter samples had the stronger trends ($\beta_{DCM} = 22.6\pm5.53$, $r= .593$, p-value < 0.001; $\beta_{175m} = 23.9\pm7.36$, $r= .503$, p-value = 0.003) while 15 meter samples showed no significant relationship ($\beta_{15m} = 3.4\pm12.9$, $r= .049$, p-value = 0.792). This meant that metabolites contributed a fixed fraction of carbon to the particulate pool with our 53 quantified metabolites representing between 1.5% and 5% of the carbon in the system (Figure 2) [Benedetto: I find this very informative. It would be great to get a ballpark sense of the percentages including the unknown metabolites, but I understand that it might be too uncertain considering the required assumptions on molecular formulas and concentration. I would like to know your wild guess, though]. Particulate nitrogen was tightly correlated with particulate carbon ($r=.933$) but the fraction of particulate nitrogen represented by the quantified metabolites was slightly higher across the transect with contributions typically between 2% and 6%.

@@

@@

targ_gp_w_sla_frac <- egg::ggarrange(barplot_targ_gp, pc_frac_metab_gp, draw = FALSE)

ggsave("targ_gp_w_sla_frac.png", plot = targ_gp_w_sla_frac, device = "png", 
       path = "figures", width = 6.875, height = 4.5, units = "in", dpi = 300, 
       bg = "white")
```

@@

*Figure 3: Differences in known metabolite concentration across the pair of adjacent eddies in the MESO-SCOPE transect separated by depth. Top panels depict concentrations of known compounds measured, where bar height corresponds to median triplicate concentration for each metabolite with the top 9 shown and the 44 other identified metabolites summed in grey. TMAO = trimethylamine N-oxide, HO-Ile = hydroxyisoleucine, GBT = glycine betaine, DCM = deep chlorophyll maximum. Lower panels depict the fraction of total particulate carbon in the known metabolites, with the mean value of three biological triplicates in color and the raw values in black behind. Colors correspond to corrected sea level anomaly (Corr. SLA), with dark red indicating anticyclonic (positive SLA) and dark blue indicating cyclonic (negative SLA) eddy state.*

@@

### A single sample highlights a network of related metabolites

@@

Isethionate is a known osmolyte thus far found exclusively in eukaryotes [@Durham2022] and strongly associated with a few diatoms in particular [@Heal2021]. This compound was enriched in the cyclone of the DCM in both the MESO-SCOPE and HEE cruises along with its precursor taurine, validating earlier results from @Barone2022 that showed strong enrichment of eukaryotic phytoplankton at the DCM and *Pseudo-nitzschia* in particular. Taurine was also among the compounds highly enriched in the anomalous 15 meter sample from Station 11, likely along with its N-trimethylated form taurine betaine (N-trimethyltaurine). While putative in our analysis, taurine betaine has been previously reported in corals and red algae, where it probably has a destabilizing influence on proteins [@Impellizzeri1975; @Hill2017; @Bruzdziak2018]. While *Trichodesmium* are notoriously patchy [@Rouco2014; @Rouco2016], we did not see a strong match to the cultured cyanobacteria in any of the other metabolites with this pattern and the osmolyte homoserine betaine shown by @Pade2016 to be characteristic of *T. erythraeum* strain IMS101 was missing.

@@

A final pattern worth noting is the enrichment of metabolites in the urea-ornithine cycle (OUC) among the samples taken during the HEE cruise. Aspartate, citrulline, ornithine, and malic acid were all found to be more abundant in the HEE samples relative to MESO-SCOPE. The OUC is a staple of metazoan metabolism and is present in stramenopiles as well where it enables rapid response to increases in nitrogen supply [@Allen2011; @Bender2014]. The enrichment of these metabolites in the HEE cruise may indicate that there was still rapid uptake and turnover of nitrogen in the 2018 eddy pair while the 2017 eddies were less active, despite both eddy pairs being in a similar state of stable anticyclone and weakening cyclone [@Dugenne2023].

Differences between the two cruises could also be due to differences in eddy age and/or the different period of the year when sampling took place during the two cruises (March/April in 2018 instead of the June/July sampling in 2017) resulting in shifts in community composition and response. @Barone2022 documents the 2017 cruise community composition, reporting abundant Prymnesiophyceae at both 15 meters and the DCM, with other significant contributions from dinoflagellates, diatoms, and green algae. In the 2018 cruise both haptophytes and dinoflagellates again dominated the eukaryotic fraction, though surprisingly little was contributed by diatoms [@Gleich2024]. @Dugenne2023 and @Harke2021 also highlight uncontrolled differences between the two sets of eddies based on their origin and age.

@@

All statistics were run in R version 4.3.3. We used multivariate statistics provided by the vegan package [@R-vegan] to assess the overall impact of various metrics across the metabolome followed by univariate statistics to investigate individually significant responses. Non-parametric tests were used when data were unlikely to obey parametric assumptions and permutational statistics were preferred wherever possible. Multiple comparison problems were controlled where necessary using a false-discovery rate correction [@Benjamini1995]. For the analyses involving the eddy transect samples, sea level anomaly was treated as a continuous variable with the exception of the PERMANOVA, for which SLA was categorized as cyclonic if less than -5 cm and anticyclonic if greater than 5 cm, with a third category for stations in between [@Anderson2001]. Both the MESO-SCOPE eddy center and HEE data treated SLA as a categorical variable.

@@

# Supplement
## Supp fig 1: PCA axes 1-3 vs SLA by depth

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE)
library(tidyverse)
library(RaMS)
library(vegan)
library(cowplot)
library(viridis)
library(scales)
library(ggnewscale)
library(egg)
library(broom)
library(lme4)
library(openxlsx)
library(readxl)
library(ggh4x)
library(ggtext)
library(gridExtra)
library(openxlsx)

@@

This shift in metabolite abundance was largely driven by a corresponding shift in biomass. Particulate carbon and summed metabolite concentration were tightly correlated across all samples and depths ($\beta = 24.9\pm1.82$ (SE), Pearson's $r$ = .813, p-value < 0.001, Supplemental Figure 2). This overall trend was largely driven by higher values at 15 meters and the DCM than at depth, but when analyzed at each depth individually the general correspondence held (Supplemental Figure 2). The DCM and 175 meter samples had the stronger trends ($\beta_{DCM} = 22.6\pm5.53$, $r= .593$, p-value < 0.001; $\beta_{175m} = 23.9\pm7.36$, $r= .503$, p-value = 0.003) while 15 meter samples showed no significant relationship ($\beta_{15m} = 3.4\pm12.9$, $r= .049$, p-value = 0.792). This meant that metabolites contributed a relatively fixed fraction of carbon to the particulate pool with our 53 quantified metabolites representing between 1.5% and 5% of the carbon in the system (Figure 2). Particulate nitrogen was tightly correlated with particulate carbon ($r=.933$) but the fraction of particulate nitrogen represented by the quantified metabolites was slightly higher across the transect with contributions typically between 2% and 6%.

Although the untargeted pool contains more molecular features, we expect that a majority of the signal has been captured by our 53 targeted compounds. Of the total peak area that passed quality control, approximately 67% was captured by these 53 molecules. An additional 15% of the total peak area was putatively annotated as various inorganic ions for which no authentic standard was available but could be matched by mass and isotope pattern. Additionally, our targeted list accounted for 13 out of the top 20 largest molecular features by total peak area with an additional 4 features annotated as the inorganic ions for 17/20 of the largest features known. However, these calculations should be taken as estimates because peak area does not correspond directly to environmental concentration and it is entirely possible for abundant environmental compounds to have small peak areas if they ionize poorly on the mass spectrometer or degrade during the sampling and extraction process.

```{r untarg targ comparison, eval=FALSE}
untarg_peaks %>%
  select(feature, filename, norm_area) %>%
  left_join(untarg_feats %>% select(feature, compound_name)) %>%
  group_by(is.na(compound_name)) %>%
  summarise(s=sum(norm_area))

summed_labels <- untarg_peaks %>%
  filter(str_detect(feature, "MS")) %>%
  group_by(feature) %>%
  summarise(s=sum(norm_area)) %>%
  arrange(desc(s)) %>%
  left_join(untarg_feats %>% select(feature, mzmed, rtmed, compound_name, polarity))
  mutate(compound_name=ifelse(feature=="MS_NEG_0155", "Bromine79", compound_name)) %>%
  mutate(compound_name=ifelse(feature=="MS_NEG_0778", "Iodine", compound_name)) %>%
  mutate(compound_name=ifelse(feature=="MS_NEG_0161", "Sulfite", compound_name)) %>%
  mutate(compound_name=ifelse(feature=="MS_NEG_0039", "Nitrate", compound_name)) %>%
  mutate(compound_name=ifelse(feature=="MS_POS_1122", "TMAO [2M+H]", compound_name))

summed_labels %>%
  group_by(is.na(compound_name)) %>%
  count()
summed_labels %>%
  group_by(is.na(compound_name)) %>%
  summarize(s=sum(s)/1e9)

summed_labels %>%
  filter(feature!="MS_NEG_0055") %>%
  filter(feature!="MS_NEG_0068") %>%
  head(20) %>%
  print(n=Inf)

msdata <- grabMSdata(list.files("../tmzMLs/neg/", full.names=TRUE, pattern="180821_Poo"))
v <- msdata$MS1[mz%between%pmppm(146.0451, 5)]
qplotMS1data(v)
```

@@

pcbarplot <- frac_pc_in_metab_df %>% 
  distinct(station, depth, PC_um) %>%
  ggplot(aes(x=station, y=PC_um)) +
  geom_col(fill="grey30") +
  facet_grid(~depth, scales="free") +
  labs(y="Particulate\ncarbon (μM)") +
  scale_x_discrete(breaks=as.character(seq(4, 14, 2))) +
  scale_y_continuous(expand = c(0.005, 0.1)) +
  theme_bw() +
  theme(strip.background = element_blank(), strip.text = element_blank(),
        axis.text.x = element_blank(), axis.title.x = element_blank(),
        plot.margin = margin(0,1,0,0), plot.background = element_rect(fill="white"))

@@

# targ_gp_w_sla_frac <- egg::ggarrange(barplot_targ_gp, pc_frac_metab_gp, draw = FALSE)
# ggsave("targ_gp_w_sla_frac.png", plot = targ_gp_w_sla_frac, device = "png", 
#        path = "figures", width = 6.875, height = 4.5, units = "in", dpi = 300, 
#        bg = "white")
targ_gp_w_sla_frac <- egg::ggarrange(barplot_targ_gp, pcbarplot, pc_frac_metab_gp,
                                     draw = FALSE, heights = c(0.4, 0.2, 0.4))
ggsave("targ_gp_w_sla_frac.png", plot = targ_gp_w_sla_frac, device = "png",
       path = "figures", width = 6.875, height = 5.5, units = "in", dpi = 300,
       bg = "white")
```

@@

*Figure 3: Differences in known metabolite concentration across the pair of adjacent eddies in the MESO-SCOPE transect separated by depth. Top panels depict concentrations of known compounds measured, where bar height corresponds to median triplicate concentration for each metabolite with the top 9 shown and the 44 other identified metabolites summed in grey. TMAO = trimethylamine N-oxide, HO-Ile = hydroxyisoleucine, GBT = glycine betaine, DCM = deep chlorophyll maximum. Center panels show the corresponding measurements of particulate carbon (PC), while the lower panels depict the fraction of total particulate carbon in the known metabolites. The mean value of three biological triplicates is shown in color and the raw values in black behind. Colors correspond to corrected sea level anomaly (Corr. SLA), with dark red indicating anticyclonic (positive SLA) and dark blue indicating cyclonic (negative SLA) eddy state.*

@@

@@

Isethionate is a known osmolyte thus far found exclusively in eukaryotes [@Durham2022] and strongly associated with a few diatoms in particular [@Heal2021]. This compound was enriched in the cyclone of the DCM in both the MESO-SCOPE and HEE cruises along with its precursor taurine, validating earlier results from @Barone2022 that showed strong enrichment of eukaryotic phytoplankton at the DCM and *Pseudo-nitzschia* in particular.

@@

Differences between the two cruises could be due to differences in eddy age and/or the different period of the year when sampling took place during the two cruises (March/April in 2018 instead of the June/July sampling in 2017) resulting in shifts in community composition and response. @Barone2022 documents the 2017 cruise community composition, reporting abundant Prymnesiophyceae at both 15 meters and the DCM, with other significant contributions from dinoflagellates, diatoms, and green algae. In the 2018 cruise both haptophytes and dinoflagellates again dominated the eukaryotic fraction, though surprisingly little was contributed by diatoms [@Gleich2024]. @Dugenne2023 and @Harke2021 also highlight uncontrolled differences between the two sets of eddies based on their origin and age.

@@

All statistics were run in R version 4.3.3. We used multivariate statistics provided by the vegan package [@R-vegan] to assess the overall impact of various metrics across the metabolome followed by univariate statistics to investigate individually significant responses. Non-parametric tests were used when data were unlikely to obey parametric assumptions and permutational statistics were preferred wherever possible. Multiple comparison problems were controlled where necessary using a false-discovery rate correction [@Benjamini1995]. For the analyses involving the eddy transect samples, sea level anomaly was treated as a continuous variable with the exception of the PERMANOVA, for which SLA was categorized as cyclonic if less than -5 cm and anticyclonic if greater than 5 cm, with a third category for stations in between [@Anderson2001]. Both the MESO-SCOPE eddy center and HEE analyses treated SLA as a categorical variable.

@@

# Supplement
## Supplemental methods: Trichodesmium bloom in MS11C215mA

## Supp fig 1: PCA axes 1-3 vs SLA by depth
